---
title: "CA"
date: "2025-04-24"
output:
  html_document:
    theme: simplex
    toc: yes
    toc_float: yes
    progress: hide
    code_folding: hide
    code_download: yes
  params:
  include_warnings: false  # Add this line to suppress warnings
  
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('/Users/jiatong/Desktop/MUSA-Praticum-Jiatong/State_agg_Rmd')
rm(list = ls())
library(httr)
library(jsonlite)
library(sf)
library(knitr)
library(ggplot2)

library(tidycensus)
library(dplyr)
library(sf)
library(tidyverse)
library(readxl)

library(purrr)
library(stringr)
```


# File import - data pre-processing


## Utility data

there have 3 sheet in this excel file. 
- sheet 1: most of zip code range locate in south california
- sheet 2: most of the zip code range locate in northern california
- sheet 3: containing zip code of southern and north california 

```{r sheet, warning=FALSE}
file_path <- "State-Specific Data/CA Zip Code ALK Data.xlsx"

sheet1 <- read_excel(file_path, sheet = 1)
sheet2 <- read_excel(file_path, sheet = 2)
sheet3 <- read_excel(file_path, sheet = 3)
```
## HUD zip code crosswalk data

The following code chunk that imports HUD zip code crosswalk data.

```{r CA crosswalk}

key <- "eyJ0eXAiOiJKV1QiLCJhbGciOiJSUzI1NiJ9.eyJhdWQiOiI2IiwianRpIjoiODk1ZWRlZDM4MjgwOTIyZDFlY2Q2MmYxMjMwODcwMDYwNGYxNTZiYzE5NWEzYmJlMmYwMGRhMmY5MDQ0MzI3N2YzM2E3NGE0M2Q2ZjY2ZmIiLCJpYXQiOjE3NDM3MDY1NDkuNzYyMzcxLCJuYmYiOjE3NDM3MDY1NDkuNzYyMzc0LCJleHAiOjIwNTkyMzkzNDkuNzQ0NzQ1LCJzdWIiOiI5NDUxMSIsInNjb3BlcyI6W119.aw4CshBNx4n3m-y1jjt1OrNh28BaRr3mNBH51OptRk_FCNZk8Su68E06d6tZ7UViafXsSeI1-OzazhGqnMzSJA"
url <- "https://www.huduser.gov/hudapi/public/usps"

# Send the GET request
response_CA <- httr::GET(url, query = list(type = 2, query = "CA"), 
                         add_headers(Authorization = paste("Bearer", key)))

# Check for errors
if (httr::http_error(response_CA)) {
  stop("Error: ", httr::http_status(response_CA)$message)
}

# Access the output as text
outputCA <- httr::content(response_CA, as = "text", encoding = "UTF-8")

# Parse JSON content from the text
output_json_CA <- fromJSON(outputCA)

# Convert to data frame (check if it's nested first)
CA_crosswalk <- as.data.frame(output_json_CA)

# View the first few rows
kable(head(CA_crosswalk))


```

## Census for GEOID (county name)

> Since HUD file does not have County name, but it contains `data.results.geoid`, which is county of GEOID.

```{r census, include=FALSE}

# Set your Census API key (only once per session)
# census_api_key("YOUR_API_KEY", install = TRUE)

# Get WA county geometries
CA_counties <- get_acs(
  geography = "county",
  state = "CA",
  variables = "B01003_001",  # total population (placeholder variable)
  geometry = TRUE,
  year = 2021
) %>%
  select(GEOID, NAME, geometry)

kable(head(CA_counties))

```

```{r, include=FALSE}
#st_write(CA_counties, "/Users/jiatong/Desktop/MUSA-practicum/WA/data/CA_counties.geojson", driver = "GeoJSON")
```

## Merging census and HUD crosswalk data to get the county name

``GEOID``in CA_counties  = ``data.results.geoid`` in CA_crosswalk
Therefore, here merges the GEOID to access the county name

```{r}

# Convert both columns to character if they are not already
CA_crosswalk$data.results.geoid <- as.character(CA_crosswalk$data.results.geoid)
CA_counties$GEOID <- as.character(CA_counties$GEOID)

# Check column names
#colnames(OR_counties)
# colnames(OR_data)

# Merge by GEOID and County_FIPS
merged_data_CA <- merge(CA_counties, CA_crosswalk, by.x = "GEOID", by.y = "data.results.geoid", all.x = TRUE) %>% st_drop_geometry()

# Display the merged data
kable(head(merged_data_CA))
```

```{r}


merged_data_CA_1 <- merged_data_CA %>% 
  select(-data.year, -data.input, -data.quarter, -data.crosswalk_type, -data.results.city, -data.results.state)


merged_data_CA_1 <- st_drop_geometry(merged_data_CA_1)

```


## removing the duplicating zip code

I will following Wilson and Din (2018) suggestion, remove the duplicating zip code that determines the county base on the largest ratio.

```{r filtered_data_CA}
filtered_data_CA <- merged_data_CA %>%
  group_by(data.results.zip) %>%
  filter(data.results.tot_ratio == max(data.results.tot_ratio)) %>%
  ungroup() %>% 
  select(-data.year, -data.quarter, -data.results.state, -data.crosswalk_type)

# View the filtered data
kable(head(filtered_data_CA))
```

# Merging utility data with zip code data
There has 3 sheet in the california disconnection rate excel file.

## Sheet 1: SoCal- LADWP
Most of the zip code located in southern California by utility **LADWP**

only showing 2 columns without county, which also without zip code.
```{r}
# Convert both columns to character if they are not already
filtered_data_CA$data.results.zip <- as.character(filtered_data_CA$data.results.zip)
sheet1$`Zip Code` <- as.character(sheet1$`Zip Code`)

merged_utility_LADWP <- merge(sheet1, filtered_data_CA, by.x = "Zip Code", by.y = "data.results.zip", all.x = TRUE) 


```

```{r}

merged_utility_LADWP <- merged_utility_LADWP %>%
  rename(
    year = Year ,
    utility_name = `Utility Name`,
    month = Month )

```


```{r}
merged_utility_LADWP_1 <- merged_utility_LADWP %>%   
  mutate(      
    est_disconnections = number_of_disconnections * data.results.tot_ratio,
    difference = number_of_disconnections - est_disconnections,
    PCT_DIF = ifelse(number_of_disconnections == 0, NA,
                     100 * (est_disconnections - number_of_disconnections) / number_of_disconnections),
    mae = abs(difference),  
    mape = case_when(
      number_of_disconnections == 0 & est_disconnections == 0 ~ 0,  
      number_of_disconnections == 0 ~ NA_real_,                   
      TRUE ~ abs(difference) / number_of_disconnections * 100     
    ))
  
```


## Sheet 2: NoCal-PGE
Most of the zip code located in northern california

```{r}
sheet2$zip_code <- as.character(sheet2$zip_code)
merged_utility_PGE <- merge(sheet2, filtered_data_CA, by.x = "zip_code", by.y = "data.results.zip", all.x = TRUE)

```


```{r, include=FALSE}
merged_utility_PGE %>% filter(is.na(GEOID)) %>% distinct(zip_code)

PGE_nd_zip_code <- merged_utility_PGE %>% filter(is.na(GEOID)) 

PGE_nd_zip_code %>% distinct(zip_code)
```

In this "merged_utility_PGE", found that wire zip code value. 

```{r}
head(merged_utility_PGE %>% filter(is.na(NAME)) %>% distinct(zip_code)) 
```

#### fill County name if pair of zip codes are same zip code

using hud data to define the zip code if pair of zip codes are same in the hud data.

```{r, include=FALSE}
zip_dict <- setNames(merged_data_CA$NAME, merged_data_CA$data.results.zip)

get_same_county <- function(zip_string) {
  matched_zips <- str_extract_all(zip_string, "\\d{5}")[[1]]
  matched_counties <- unique(na.omit(zip_dict[matched_zips]))
    if (length(matched_counties) == 1) {
    return(matched_counties)
  } else {
    return(NA_character_)
  }
}

merged_utility_PGE <- merged_utility_PGE %>%
  mutate(
        #new_name = map_chr(zip_code, get_same_county),
        NAME = map_chr(zip_code, get_same_county)
        #data.results.tot_ratio = if_else(!is.na(new_name), 1, data.results.tot_ratio)
) 

merged_utility_PGE %>% filter(is.na(NAME)) %>% distinct(zip_code)

```


#### fill County name if pair of zip codes are not same zip code

```{r}
merged_utility_PGE <- merged_utility_PGE %>%
  mutate(NAME = case_when(
    zip_code == "93212/93219" ~ "Tulare County, California",
    zip_code == "93282" ~ "Tulare County, California",
    zip_code == "93428/93429" ~ "San Luis Obispo County, California",
    zip_code == "93607" ~ "Fresno County, California",
    zip_code == "93667/93670" ~ "Fresno County, California",
    zip_code == "93668/93670" ~ "Fresno County, California",
    zip_code == "94021 / 94022" ~ "San Mateo County, California",
    zip_code == "94515/94516" ~ "Napa County, California",
    zip_code == "94534/94535" ~ "Solano County, California",
    zip_code == "94536/94535" ~ "Alameda County, California",
    zip_code == "94568/94569" ~ "Alameda County, California",
    zip_code == "94801/94720" ~ "Contra Costa County, California",
    zip_code == "94971/94972" ~ "Marin County, California",
    zip_code == "95002/94972" ~ "Santa Clara County, California",
    zip_code == "95226 / 95227" ~ "Calaveras County, California",
    zip_code == "95234 / 95236" ~ "San Joaquin County, California",
    zip_code == "95250" ~ "Calaveras County, California",
    zip_code == "95252/95253" ~ "Calaveras County, California",
    zip_code == "95253 / 95254" ~ "Calaveras County, California",
    zip_code == "95322/95325" ~ "Merced County, California",
    zip_code == "95383/95385" ~ "Tuolumne County, California",
    zip_code == "95428/95430" ~ "Mendocino County, California",
    zip_code == "95430 / 95432" ~ "Mendocino County, California",
    zip_code == "95476/95480" ~ "Sonoma County, California",
    zip_code == "95482/95480" ~ "Mendocino County, California",
    zip_code == "95552 / 95553" ~ "Humboldt County, California",
    zip_code == "95585/95587" ~ "Mendocino County, California",
    zip_code == "95585/95595" ~ "Mendocino County, California",
    zip_code == "95589/95595" ~ "Mendocino County, California",
    zip_code == "95602/95595" ~ "Placer County, California",
    zip_code == "95616/95615" ~ "Yolo County, California",
    zip_code == "95643" ~ "Amador County, California",
    zip_code == "95651/95654" ~ "El Dorado County, California",
    zip_code == "95653/95654" ~ "Yolo County, California",
    zip_code == "95677/95679" ~ "Placer County, California",
    zip_code == "95677/95679/95680" ~ "Placer County, California",
    zip_code == "95681/95679" ~ "Placer County, California",
    zip_code == "95695/95699" ~ "Yolo County, California",
    zip_code == "95697/95699" ~ "Yolo County, California",
    zip_code == "95698 / 95701" ~ "Yolo County, California",
    zip_code == "95901/95910" ~ "Yuba County, California",
    zip_code == "95914/95915" ~ "Butte County, California",
    zip_code == "95982/95984" ~ "Sutter County, California",
    zip_code == "96009/96011" ~ "Shasta County, California",
    zip_code == "96028/96029" ~ "Siskiyou County, California",
    zip_code == "96029 / 96033" ~ "Siskiyou County, California",
    zip_code == "96065/96068" ~ "Siskiyou County, California",
    zip_code == "96073/96074" ~ "Tehama County, California",
    zip_code == "96075/96076/96078" ~ "Tehama County, California",
    zip_code == "96076 / 96078" ~ "Tehama County, California",
    zip_code == "95915 / 95918" ~ "Plumas County, California",
    zip_code == "95943/95944" ~ "Glenn County, California",
    zip_code == "95944 / 95945" ~ "Plumas County, California",
    zip_code == "95957/95958" ~ "Colusa County, California",
    zip_code == "95971/95972" ~ "Plumas County, California",
    zip_code == "95972 / 95973" ~ "Butte County, California",
    zip_code == "95973/95972" ~ "Butte County, California",
    zip_code == "95982/95981" ~ "Sutter County, California",
    TRUE ~ NAME
  )) 

```

```{r, include=FALSE}
merged_utility_PGE %>% filter(is.na(NAME)) %>% distinct(zip_code)

merged_utility_PGE <- merged_utility_PGE %>%
  mutate(data.results.tot_ratio = if_else(is.na(data.results.tot_ratio), 1, data.results.tot_ratio))

```


## Sheet 3: Cal: SCE

```{r}
sheet3$zip_code <- as.character(sheet3$zip_code)
merged_utility_SCE <- merge(sheet3, filtered_data_CA, by.x = "zip_code", by.y = "data.results.zip", all.x = TRUE)

merged_utility_SCE %>% distinct(utility_name)

```

Since HUD does not provide all the county name, I need to fill the county name by search from the internet.
I have create a new column for filling all zip code county name. 
- 1. I need to make sure a format are same, which need to adjust the format
- 2.  those zip code that are not in the HUD dataset, resulting NA value in``data.results.tot_ratio`` . To adjust it, I fill the 1 in `data.results.tot_ratio``that represents all zip code are fully in the cerntain county.

--- 
-   If ZIP code is valid ➔ 
    (example: `"Baker County, Maryland"`)

-   If ZIP code is invalid ➔ `"Invalid"`

    I checked using the [ZIP code lookup](https://www.unitedstateszipcodes.org) resources.
    

```{r}
merged_utility_SCE <- merged_utility_SCE %>%
  mutate(
    NAME = case_when(
      zip_code == 93216 ~ "Kern County, California",
      zip_code == 93519 ~ "Kern County, California",
      zip_code == 93542 ~ "Inyo County, California",
      zip_code == 93558 ~ "Inyo County, California",
      TRUE ~ NAME  # keep existing NAME if ZIP not matched
    ),
    data.results.tot_ratio = ifelse(
      zip_code %in% c(93216, 93519, 93542, 93558),
      1,
      data.results.tot_ratio
    )
  )

```


```{r}
merged_utility_SCE_1 <- merged_utility_SCE %>% 
  mutate(      
    est_disconnections = number_of_disconnections * data.results.tot_ratio,
    difference = number_of_disconnections - est_disconnections,
    PCT_DIF = ifelse(number_of_disconnections == 0, NA,
                     100 * (est_disconnections - number_of_disconnections) / number_of_disconnections),
    mae = abs(difference),  
    mape = case_when(
      number_of_disconnections == 0 & est_disconnections == 0 ~ 0,  
      number_of_disconnections == 0 ~ NA_real_,                   
      TRUE ~ abs(difference) / number_of_disconnections * 100     
    )) 
```


# Aggregate data from zip code to county level

The code aggregates utility disconnection data at the **county level** in Maryland State by grouping the dataset `merged_utility_MD` based on `NAME` (county), `year`, `month`, and `utility_name`.
The resulting summary table `WA_county_agg` includes:

-   **`total_disconnections`**: The total number of reported utility disconnections in each group.

-   **`est_disconnections`**: The estimated number of disconnections, adjusted using a weighting factor (`data.results.tot_ratio`), which accounts for partial ZIP-to-county mappings.

-   **`avg_disconnection_rate`**: The average disconnection count per record in each zip code

-   **`difference`**: The numerical difference between the **`total_disconnections`** and **`est_disconnections`** totals, which can help assess the impact of the weighting method.

- **`PCT_DIF`**: the precentage of **`difference`**
- **`mae`** (Mean Absolute Error): mean of absolute differences between total_disconnections and est_disconnections

- **`mape`**(Mean Absolute Percentage Error):
the percentage equivalent of mean absolute error (MAE), measuring the accuracy between total_disconnections and est_disconnections

## CA_county: sce

```{r CA_county_agg}
SCE_county_agg <- merged_utility_SCE %>%
  group_by(NAME, year, month, utility_name) %>%
  summarise(
    total_disconnections = sum(number_of_disconnections, na.rm = TRUE),
    #total_disconnections_rate = sum(disconnection_rate, na.rm = TRUE),
    est_disconnections = sum(number_of_disconnections * data.results.tot_ratio, na.rm = TRUE),
    avg_disconnection_rate = mean(number_of_disconnections, na.rm = TRUE),
    difference = sum(total_disconnections - est_disconnections, na.rm = TRUE),  
    PCT_DIF = case_when(
      sum(number_of_disconnections, na.rm = TRUE) == 0 & sum(est_disconnections, na.rm = TRUE) == 0 ~ 0,
      sum(number_of_disconnections, na.rm = TRUE) == 0 ~ NA_real_,
      TRUE ~ 100 * (sum(est_disconnections, na.rm = TRUE) - sum(number_of_disconnections, na.rm = TRUE)) / sum(number_of_disconnections, na.rm = TRUE)
    ),
    mae = abs(sum(difference, na.rm = TRUE)),
    mape = case_when(
      sum(number_of_disconnections, na.rm = TRUE) == 0 & sum(est_disconnections, na.rm = TRUE) == 0 ~ 0,
      sum(number_of_disconnections, na.rm = TRUE) == 0 ~ NA_real_,
      TRUE ~ abs(sum(difference, na.rm = TRUE)) / sum(number_of_disconnections, na.rm = TRUE) * 100
    ),
    .groups = "drop"   
  )

kable(head(SCE_county_agg))
```


## Socal: LADWP

```{r Socal_county_agg}
LADWP_county_agg <- merged_utility_LADWP %>% 
  group_by(., NAME, year, month, utility_name) %>% 
  summarise(
    total_disconnections = sum(number_of_disconnections, na.rm = TRUE),
    est_disconnections = sum(number_of_disconnections * data.results.tot_ratio, na.rm = TRUE),
    avg_disconnection_rate = mean(number_of_disconnections, na.rm = TRUE),
    difference = sum(total_disconnections - est_disconnections, na.rm = TRUE),  
    PCT_DIF = case_when(
      sum(number_of_disconnections, na.rm = TRUE) == 0 & sum(est_disconnections, na.rm = TRUE) == 0 ~ 0,
      sum(number_of_disconnections, na.rm = TRUE) == 0 ~ NA_real_,
      TRUE ~ 100 * (sum(est_disconnections, na.rm = TRUE) - sum(number_of_disconnections, na.rm = TRUE)) / sum(number_of_disconnections, na.rm = TRUE)
    ),
    mae = abs(sum(difference, na.rm = TRUE)),
    mape = case_when(
      sum(number_of_disconnections, na.rm = TRUE) == 0 & sum(est_disconnections, na.rm = TRUE) == 0 ~ 0,
      sum(number_of_disconnections, na.rm = TRUE) == 0 ~ NA_real_,
      TRUE ~ abs(sum(difference, na.rm = TRUE)) / sum(number_of_disconnections, na.rm = TRUE) * 100
    ),
    .groups = "drop"   
  )

kable(head(LADWP_county_agg))

```



## PGE_county_agg

```{r}
PGE_county_agg <- merged_utility_PGE %>% 
  group_by(., NAME, year, month, utility_name) %>% 
  summarise(
    total_disconnections = sum(number_of_disconnections, na.rm = TRUE),
    est_disconnections = sum(number_of_disconnections * data.results.tot_ratio, na.rm = TRUE),
    avg_disconnection_rate = mean(number_of_disconnections, na.rm = TRUE),
    difference = sum(total_disconnections - est_disconnections, na.rm = TRUE),  
    PCT_DIF = case_when(
      sum(number_of_disconnections, na.rm = TRUE) == 0 & sum(est_disconnections, na.rm = TRUE) == 0 ~ 0,
      sum(number_of_disconnections, na.rm = TRUE) == 0 ~ NA_real_,
      TRUE ~ 100 * (sum(est_disconnections, na.rm = TRUE) - sum(number_of_disconnections, na.rm = TRUE)) / sum(number_of_disconnections, na.rm = TRUE)
    ),
    mae = abs(sum(difference, na.rm = TRUE)),
    mape = case_when(
      sum(number_of_disconnections, na.rm = TRUE) == 0 & sum(est_disconnections, na.rm = TRUE) == 0 ~ 0,
      sum(number_of_disconnections, na.rm = TRUE) == 0 ~ NA_real_,
      TRUE ~ abs(sum(difference, na.rm = TRUE)) / sum(number_of_disconnections, na.rm = TRUE) * 100
    ),
    .groups = "drop"   
  )

kable(head(PGE_county_agg))
```

```{r, include=FALSE}
names(SCE_county_agg)
names(LADWP_county_agg)
names(PGE_county_agg)

```

> Combining those 3 excel `county_agg` together.

```{r}
CA_county_agg <- rbind(SCE_county_agg, LADWP_county_agg, PGE_county_agg) 


#CA_county_agg %>% group_by(NAME, year, month, utility_name)
```


```{r}
CA_county_agg_geo_full <- CA_county_agg %>%
  left_join(CA_counties, by = "NAME") %>%
  st_as_sf()

#st_write(CA_county_agg_geo_full, "CA_county_agg_summary.geojson", driver = "GeoJSON")
# write_csv(CA_county_agg, "CA_county_agg_summary.csv")

```

> The followng plot visulazie Cumulative Total Disconnections in Disconnection Estimates (2018–2024)

```{r}

# Step 1: group by NAME + year，summary total_disconnections
CA_county_agg_summary <- CA_county_agg %>%
  group_by(NAME, year) %>%
  summarise(
    total_disconnections = sum(total_disconnections, na.rm = TRUE),
    .groups = "drop"
  )

# Step 2: creating grid
Cal_county_year_grid <- expand_grid(
  NAME = unique(CA_counties$NAME),
  year = unique(CA_county_agg$year)
)

# Step 3: Left join
Cal_county_agg_geo <- Cal_county_year_grid %>%
  left_join(CA_counties, by = "NAME") %>%
  left_join(CA_county_agg_summary, by = c("NAME", "year")) %>%
  st_as_sf()

# Step 4: 
ggplot(Cal_county_agg_geo) +
  geom_sf(aes(fill = total_disconnections), color = "white") +
  facet_wrap(~ year) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt", na.value = "grey80") +
  labs(
    title = "Cumulative Total Disconnections in Disconnection Estimates (2013–2020)",
    fill = "Total Disconnections",
    caption = "Source: Aggregated ZIP-to-County Estimates\nUnit: Disconnection Count Difference"
  ) +
  theme_minimal() +
    theme(
    axis.text = element_blank(),   
    axis.ticks = element_blank(),  
    panel.grid = element_blank()  
  )
```

# Vadilating 

## comparing **total_disconnections and est_disconnections**

In previous aggregating step,  aggregating the disconnection data by county, utility company, year, and month.

- total_disconnections: the unweighted sum of disconnections reported within each group.

- est_disconnections: the weighted sum, using the ratio to estimate each ZIP’s contribution to the county.

### Socal_county_agg

> To evaluate how much the weighting affected the total count:
>
 -   I summed the **unweighted totals** (`total_disconnections`) across all counties, resulting in **139552 disconnections**.

 -   I summed the **weighted totals**(`est_disconnections`), yielding **139473.818920778 disconnections**.

 -   The **absolute difference** between the two totals was **78.1810792222672 disconnections**, indicating the net shift caused by applying HUD’s weighting.
```{r}
print(paste("rows - unweighted total(sum of number_of_disconnections):", sum(CA_county_agg$total_disconnections)))
print(paste("rows - weighted total(est_disconnection) :", sum(CA_county_agg$est_disconnections)))
print(paste("Difference (impact of weighting):", sum(CA_county_agg$difference)))
```

> **Relative difference**
>
> To contextualize this shift, I calculated the **relative difference** as a percentage of the unweighted total:

```{r}

relative_diff_socal <- 
  sum(CA_county_agg$difference) / sum(CA_county_agg$total_disconnections)* 100

print(paste("Relative Difference (%):", round(relative_diff_socal, 2), "%"))
```


The **0.4% relative difference** indicates that applying the ZIP-to-county weighting introduces **minimal deviation** from the raw totals.




## Evaluation Metrics

In this section, I'm gonna to **Calculate Evaluation Metrics**, which quantify the estimation error across all grouped observations (by county, month, and utility):

MAE (Mean Absolute Error): mean of absolute differences between total_disconnections and est_disconnections
- MAE is not a normalized metric and is dependent on the scale of the data. This means that an MAE value of 10 might be considered good for predicting millions, but poor for predicting values in the range of 2 digits. 


$$
\text{Mean Absolute Error} = \frac{1}{n} \sum_{i=1}^{n} \left| \text{total_disconnections}_i - \text{est_disconnections}_i \right|
$$

-   MAE value range is 0 from infinity, therefore, less than 1 indicting good accuracy.

MAPE (Mean Absolute Percentage Error):

the percentage equivalent of mean absolute error (MAE), measuring the accuracy between total_disconnections and est_disconnections

$$
\text{Mean Percentage Error} = \frac{1}{n} \sum_{i=1}^{n} \left( \frac{\text{total_disconnections} - \text{est_disconnections}_i}{\text{total_disconnections}_i} \right) \times 100
$$

> [Montaño, Juan, and Palmer et al](https://www.researchgate.net/publication/257812432_Using_the_R-MAPE_index_as_a_resistant_measure_of_forecast_accuracy). suggest that a MAPE below 20% indicates good accuracy.

> In general, the MAPE across California State remains under this 20% threshold, and 1.33% MAPE value represents high accuracy across California in county level.


```{r}
mae_CA <- mean(CA_county_agg$mae, na.rm = TRUE)
mape_CA <- mean(CA_county_agg$mape, na.rm = TRUE)  # percentage
rmse_CA <- sqrt(mean((CA_county_agg$total_disconnections - CA_county_agg$est_disconnections)^2, na.rm = TRUE))

print(paste("MAE:", round(mae_CA, 2)))
print(paste("MAPE (%):", round(mape_CA, 2)))

```







```{r, include=FALSE}
# Create a summarized table: actual vs estimated
county_comparison_CA <- CA_county_agg %>%
  group_by(NAME) %>%
  summarise(
    actual_total = sum(total_disconnections, na.rm = TRUE),
    estimated_total = sum(est_disconnections, na.rm = TRUE)
  ) %>%
  arrange(desc(actual_total))  # optional: sort by actual



# Reshape for visualization
county_long_CA <- county_comparison_CA %>%
  pivot_longer(cols = c("actual_total", "estimated_total"),
               names_to = "type", values_to = "value")

# Plot
ggplot(county_long_CA, aes(x = reorder(NAME, -value), y = value, fill = type)) +
  geom_col(position = "dodge") +
  labs(title = "Actual vs Estimated Utility Disconnections by County",
       x = "County", y = "Number of Disconnections") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


> The following plot visualizes the MAPE of disconnection estimates from 2019 to 2024. Overall, the plot shows that MAPE values are generally below 20% for each county, indicating good accuracy.

```{r cal mape map}

# Step 1: make sure each year have County
county_year_grid_CA <- expand_grid(
  NAME = unique(CA_counties$NAME),
  year = unique(CA_county_agg$year)
)

# Step 2: summary error summary
county_error_summary_CA <- CA_county_agg %>%
  group_by(NAME, year) %>%
  summarise(
    total_abs_error = sum(mae, na.rm = TRUE),
    total_disconnections = sum(total_disconnections, na.rm = TRUE),
    total_est_disconnections = sum(est_disconnections, na.rm = TRUE),
    mape = (total_abs_error / total_disconnections) * 100,  # Percent form
    .groups = "drop"
  )

# Step 3: using grid left_join，let no data be NA value
county_error_complete_CA <- county_year_grid_CA %>%
  left_join(county_error_summary_CA, by = c("NAME", "year"))

# Step 4:  join spatial
county_error_map_CA <- CA_counties %>%
  left_join(county_error_complete_CA, by = "NAME")  

# Step 5: 
ggplot(county_error_map_CA) +
  geom_sf(aes(fill = mape), color = "white") +
  facet_wrap(~ year) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt", na.value = "grey80") +
  labs(
    title = "Cumulative MAPE in Disconnection Estimates",
    fill = "MAPE",
    caption = "Source: Aggregated ZIP-to-County Estimates\nUnit: Disconnection Count Difference"
  ) +
  theme_minimal() +
      theme(
    axis.text = element_blank(),   
    axis.ticks = element_blank(),  
    panel.grid = element_blank()  
  )
```





# Limitation

-  In this study, when aggregating utility data from ZIP code to county levels, it is not feasible to directly validate the accuracy of the estimates.

-   As noted by Din and Wilson (2020), when aggregating data, it is unlikely to have point-level information available; otherwise, there would be no need to perform crosswalk processing.

-   Although direct validation is not possible, examining differences between estimated and actual counts may reveal additional patterns of utility disconnection rate, which can be further explored through more detailed spatial analysis.

<!-- -->

#### MAPE > 20% and MAE > 1 within the zip code level data

--- MAPE: the percentage equivalent of mean absolute error (MAE), measuring the accuracy between ``total_disconnections`` and ``est_disconnections``

--- mean of absolute differences between ``total_disconnections`` and ``est_disconnections``

MAPE measures percentage error; MAE measures the average absolute difference between actual and estimated disconnections.

> In the following analysis, 79 out of 13,730 ZIP code-level records show a MAPE greater than 20% and a MAE greater than 1.

-- Some small disconnection counts can lead to a large percentage difference (high MAPE). Therefore, we also examine the MAE:
a MAE less than 1 indicates a high level of accuracy despite a seemingly large MAPE.

-- As we lack point-level data, a precise one-to-one match between ``total_disconnections`` and ``est_disconnections`` values at the ZIP code level is not feasible.




> Among 2506  ZIP code-level records, only 20 records (0.79%) showed a MAPE greater than 20% and a MAE greater than 1.



```{r}
kable(head(
  CA_county_agg  %>% filter(mape > 20 & mae > 1))
  )

nrow(CA_county_agg  %>% filter(mape > 20 & mae > 1))

```
> **Given the absence of point-level disconnection data, this very low error rate demonstrates strong model accuracy at the ZIP code aggregation level.**


# Reference

Din, A., & Wilson, R.
(2020).
Crosswalking ZIP Codes to Census Geographies: Geoprocessing the U.S.
Department of Housing & Urban Development’s ZIP Code Crosswalk Files. 
*Cityscape: A Journal of Policy Development and Research, 22*(1), 293–298.
U.S.
Department of Housing and Urban Development. 
<https://www.huduser.gov/portal/periodicals/cityscpe/vol22num1/ch12.pdf>​

Wilson, R., & Din, A.
(2018).
Understanding and enhancing the U.S.
Department of Housing and Urban Development’s ZIP Code Crosswalk Files. 
*Cityscape: A Journal of Policy Development and Research, 20*(2), 277–294. 
<https://www.huduser.gov/portal/periodicals/cityscpe/vol20num2/ch16.pdf>​

Din, A.
(2021).
New Data Fields for HUD Aggregated USPS Administrative Data on Address Vacancies.
*Cityscape*, *23*(3), 283–294.
<https://www.jstor.org/stable/48636236>

Montaño, Juan & Palmer, Alfonso & Sesé, Albert & Cajal, Berta.
(2013).
Using the R-MAPE index as a resistant measure of forecast accuracy.
Psicothema.
25.
500-506.
10.7334/psicothema2013.23.

-   <https://www.sciencedirect.com/science/article/pii/S1051200419301289#bbr0610>

-   <https://www.vexpower.com/brief/mean-absolute-percentage-error>
